---
title: "Big 5 Leagues"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: cosmo
    social: [ "twitter", "linkedin", "menu" ]
---

<style>                     
.navbar {
  background-color:black;
  border-color:white;
}
.navbar-brand {
color:white!important;
}
</style>   

<style type="text/css">

.chart-title {  /* chart_title  */
   font-size: 15px

</style>


```{r global, include = FALSE}
library(flexdashboard)
library(shiny)
library(reactable)
library(reactablefmtr)
library(worldfootballR)
library(tidyverse)
library(rvest)
library(teamcolors) # NFL team colors and logos
library(extrafont)
library(gt)
library(gtExtras)
library(ggimage)
library(animation)
library(DBI)
library(RSQLite)
library(glue)
library(ggtext)
library(scales)


## Setup steps

team_mapping <- read_csv("https://raw.githubusercontent.com/steodose/Club-Soccer-Forecasts/main/team_mapping.csv")

# fix Nottingham Forest in 538 team mapping data
team_mapping2 <- team_mapping %>%
    mutate(team_538 = case_when(
        team == "Nott'm Forest" ~ "Nottingham Forest",
        TRUE ~ team_538
    ))

# Optional but makes R prefer not to display numbers in scientific notation
options(scipen = 9999)

# Custom ggplot theme (inspired by Owen Phillips at the F5 substack blog)
theme_custom <- function () { 
  theme_minimal(base_size=11, base_family="Chivo") %+rbundace% 
    theme(
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = 'transparent', color = "transparent")
    )
}

# Define 538 table theme for Reactable table(s) below
theme_538 <- function() {
    reactable::reactableTheme(
        searchInputStyle = list(width = "31%", backgroundColor = "#F9F9F9"),
        style = list(
            fontFamily = "Chivo"
        ),
        headerStyle = list(
            "&:hover[aria-sort]" = list(
                background = "hsl(0, 0%, 80%)"),
            "&[aria-sort='ascending'], &[aria-sort='descending']" = list(
                background = "#555",
                color = "#FFF"
            ),
            borderColor = "#333"
        ),
        borderColor = "#CDCDCD"
    )
}


##### Custom gt table themes for graphics. Inspired by Tom Mock's excellent blog posts #####
gt_theme_538 <- function(data,...) {
    data %>%
        # Add team logos w/ web_image
        text_transform(
            locations = cells_body(
                columns = flag_logo
            ),
            fn = function(x) {
                web_image(
                    url = x,
                    height = 25
                )
            }
        ) %>%
        # Relabel columns
        cols_label(
            flag_logo = ""
        ) %>%
        opt_all_caps()  %>%
        opt_table_font(
            font = list(
                google_font("Chivo"),
                default_fonts()
            )
        ) %>%
        tab_style(
            style = cell_borders(
                sides = "bottom", color = "transparent", weight = px(2)
            ),
            locations = cells_body(
                columns = TRUE,
                # This is a relatively sneaky way of changing the bottom border
                # Regardless of data size
                rows = nrow(data$`_data`)
            )
        )  %>% 
        tab_options(
            column_labels.background.color = "white",
            table.border.top.width = px(3),
            table.border.top.color = "transparent",
            table.border.bottom.color = "transparent",
            table.border.bottom.width = px(3),
            column_labels.border.top.width = px(3),
            column_labels.border.top.color = "transparent",
            column_labels.border.bottom.width = px(3),
            column_labels.border.bottom.color = "black",
            data_row.padding = px(3),
            source_notes.font.size = 12,
            table.font.size = 16,
            heading.align = "left",
            ...
        ) 
}


## Load data

matches <- read_csv("https://projects.fivethirtyeight.com/soccer-api/club/spi_matches_latest.csv") %>%
  filter(league == 'Barclays Premier League' | league == 'German Bundesliga' | league == 'Spanish Primera Division' |  league == 'French Ligue 1' | league == 'Italy Serie A',
           season == 2022)



```

Inputs {.sidebar}
-----------------------------------------------------------------------

``` {r}

 selectInput("leagueInput", label = h3("Select League"), 
    choices = unique(matches$league),
    selected="Barclays Premier League", multiple = FALSE)

```


Row
-----------------------------------------------------------------------

### Country {.value-box data-width=50}

```{r}

# make dataframe with country names
country_leagues <- data.frame(country = c("England", "Germany", "Spain", "France", "Italy"),
                        league = c("Barclays Premier League", "German Bundesliga", "Spanish Primera Division", "French Ligue 1", "Italy Serie A")
)

country<- reactive({
  country_leagues %>% 
  filter(league == input$leagueInput) %>% 
  pull(country)
})


# Put valueBox() inside renderValueBox({})
renderValueBox({
  valueBox(value = country(), 
           caption = "Country",
           color = "#013369",
           icon = "fa-globe")
})

```


### Data Updated {.value-box data-width=50}

```{r}

update_date <- reactive({
  matches %>% 
  filter(league == input$leagueInput) %>% 
    na.omit() %>%
  summarise(latest_match = max(date))
})


# Put valueBox() inside renderValueBox({})
renderValueBox({
  valueBox(value = update_date(), 
           caption = "Matches Last Updated",
           color = "#013369",
           icon = "fa-calendar")
})

```

Row {.tabset data-height=650}
-----------------------------------------------------------------------

### League Table
    
```{r}

epl_matches <- reactive({
  matches %>%
  filter(league == input$leagueInput) %>%
  mutate(result = score1 - score2,
           xresult = xg1 - xg2) %>%
    select(team1, team2, result, xresult, score1, score2, xg1, xg2) %>%
    pivot_longer(team1:team2, names_to = "home_away", values_to = "team") %>%
    mutate(
        result = ifelse(home_away == "team1", result, -result),
        xresult = ifelse(home_away == "team1", xresult, -xresult),
        win = ifelse(result == 0, 0.5, ifelse(result > 0, 1, 0))
    ) %>%
    select(team, score1, score2, xg1, xg2, win, result, xresult) %>%
    drop_na() %>%
    group_by(team) %>%
    summarise(
        Wins = length(win[win == 1]),
        Losses = length(win[win == 0]),
        Draws = length(win[win == 0.5]),
        MP = sum(Wins, Losses, Draws),
        Points = (Wins * 3) + (Draws * 1),
        win_perc = (100 * Points / (MP * 3)),
        GD = sum(result),
        xGD = sum(xresult),
        form = list(win), .groups = "drop"
    ) %>%
    left_join(team_mapping2, by = c("team" = "team_538")) %>%
    select(url_logo_espn, team, Points, MP, Wins, Draws, Losses, GD, xGD, win_perc, form) %>%
    arrange(desc(Points), desc(GD)) %>%
    ungroup() %>%
    mutate(Rank = row_number()) %>%
    relocate(Rank) %>%
    rename(Team = team) %>%
    mutate(list_data = list(c(Wins, Draws, Losses))) %>%
    gather(attr_num, list_data, c(Wins, Draws, Losses)) %>%
    group_by_at(vars(-attr_num, -list_data)) %>%
    summarise(list_data = list(list_data)) %>%
    ungroup()
})



# Make current standings table using gt package
render_gt({
epl_matches() %>%
    gt() %>%
    # Relabel columns
    cols_label(
        url_logo_espn = "",
        win_perc = "Win %",
        form = "Form"
    ) %>%
    text_transform(
        locations = cells_body(vars(url_logo_espn)),
        fn = function(x) {
            web_image(url = x,
                      height = px(30))
        }
    ) %>%
    data_color(columns = 4,
               colors = scales::col_numeric(
                   palette = c("white", "#3fc1c9"),
                   domain = NULL)
    ) %>%
    #gt_theme_538() %>%
    tab_style(
        style = cell_borders(sides = "bottom", color = "black", weight = px(1)),
        locations = cells_body(rows = 4)
    ) %>%
    tab_style(
        style = cell_borders(sides = "bottom", color = "black", weight = px(1)),
        locations = cells_body(rows = 17)
    ) %>%
    tab_style(
        style = list(
            cell_text(color = "red")
        ),
        locations = cells_body(
            columns = vars(GD),
            rows = GD <= 0
        )
    ) %>%
    tab_style(
        style = list(
            cell_text(color = "blue")
        ),
        locations = cells_body(
            columns = vars(GD),
            rows = GD > 0
        )
    ) %>%
    tab_style(
        style = list(
            cell_text(color = "red")
        ),
        locations = cells_body(
            columns = vars(xGD),
            rows = xGD <= 0
        )
    ) %>%
    tab_style(
        style = list(
            cell_text(color = "blue")
        ),
        locations = cells_body(
            columns = vars(xGD),
            rows = xGD > 0
        )
    ) %>%
    fmt_number(
        columns = vars(xGD),
        decimals = 1
    ) %>%
    cols_align(align = "left",
               columns = 1) %>%
    tab_options(column_labels.font.weight = "bold") %>%
    tab_header(title = md("**2022-23 League Table**"),
               subtitle = glue("Thru matches played as of date mentioned above.")) %>%
    tab_source_note(
        source_note = md("DATA: fivethirtyeight.com")) %>%
    gt_plt_bar_pct(column = win_perc, scaled = TRUE, fill = "navy", background = "gray") %>%
    gt_plt_winloss(form, max_wins = 30) %>%
    gt_plt_bar_stack(list_data, width = 60,
                     labels = c("  WINS  ", "  DRAWS  ", "  LOSSES  "),
                     palette= c("#ff4343", "#bfbfbf", "#0a1c2b")) %>%
    tab_footnote(
        footnote = "Share of total available points captured. Win = 3 points, Draw = 1 point, Loss = 0 points.",
        locations = cells_column_labels(vars(win_perc))
    ) %>%
    tab_footnote(
        footnote = "Expected Goals (xG) is the probability a shot will result in a goal based on the characteristics of that shot. Measures  quality of chances created.",
        locations = cells_column_labels(vars(xGD))
    )

})

```



```{css}
    .chart-shim {
      overflow: auto;
    }
```

### About

**Big 5 European Leagues App**

This dashboard looks at the current tables and statistics for the Big 5 European leagues. All data is sourced from FiveThirtyEight.com unless noted otherwise. App created using R Shiny by [Stephan Teodosescu](https://stephanteodosescu.com).