---
title: "Big 5 Leagues"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: cosmo
    social: [ "twitter", "linkedin", "menu" ]
---

<style>                     
.navbar {
  background-color:black;
  border-color:white;
}
.navbar-brand {
color:white!important;
}
</style>   

<style type="text/css">

.chart-title {  /* chart_title  */
   font-size: 15px;
}
</style>


```{r global, include = FALSE}
library(flexdashboard)
library(lubridate)
library(shiny)
library(reactable)
library(reactablefmtr)
library(htmltools)
library(tidyverse)
library(rvest)
library(extrafont)
library(gt)
library(gtExtras)
library(ggimage) 
library(glue)
library(ggtext)
library(scales)
library(svglite)


## Setup steps

# Optional but makes R prefer not to display numbers in scientific notation
options(scipen = 9999)

# Custom ggplot theme (inspired by Owen Phillips at the F5 substack blog)
theme_custom <- function () { 
  theme_minimal(base_size=11, base_family="Chivo") %+rbundace% 
    theme(
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = 'transparent', color = "transparent")
    )
}

# Define 538 table theme for Reactable table(s) below
theme_538 <- function() {
    reactable::reactableTheme(
        searchInputStyle = list(width = "31%", backgroundColor = "#F9F9F9"),
        style = list(
            fontFamily = "Chivo"
        ),
        headerStyle = list(
            "&:hover[aria-sort]" = list(
                background = "hsl(0, 0%, 80%)"),
            "&[aria-sort='ascending'], &[aria-sort='descending']" = list(
                background = "#555",
                color = "#FFF"
            ),
            borderColor = "#333"
        ),
        borderColor = "#CDCDCD"
    )
}


##### Custom gt table themes for graphics. Inspired by Tom Mock's excellent blog posts #####
gt_theme_538 <- function(data,...) {
    data %>%
        # Add team logos w/ web_image
        text_transform(
            locations = cells_body(
                columns = flag_logo
            ),
            fn = function(x) {
                web_image(
                    url = x,
                    height = 25
                )
            }
        ) %>%
        # Relabel columns
        cols_label(
            flag_logo = ""
        ) %>%
        opt_all_caps()  %>%
        opt_table_font(
            font = list(
                google_font("Chivo"),
                default_fonts()
            )
        ) %>%
        tab_style(
            style = cell_borders(
                sides = "bottom", color = "transparent", weight = px(2)
            ),
            locations = cells_body(
                columns = TRUE,
                # This is a relatively sneaky way of changing the bottom border
                # Regardless of data size
                rows = nrow(data$`_data`)
            )
        )  %>% 
        tab_options(
            column_labels.background.color = "white",
            table.border.top.width = px(3),
            table.border.top.color = "transparent",
            table.border.bottom.color = "transparent",
            table.border.bottom.width = px(3),
            column_labels.border.top.width = px(3),
            column_labels.border.top.color = "transparent",
            column_labels.border.bottom.width = px(3),
            column_labels.border.bottom.color = "black",
            data_row.padding = px(3),
            source_notes.font.size = 12,
            table.font.size = 16,
            heading.align = "left",
            ...
        ) 
}

## new function to clean/normalize matches data schema with old schema now that worldfootballR has been deprecated

clean_matches <- function(df) {

  # If the file is already in your "old" schema, keep it working
  has_old <- all(c("Competition_Name","Home","Away","HomeGoals","AwayGoals","Date") %in% names(df))

  if (!has_old) {
    # football-data.co.uk schema
    # Expect: Div, Date (dd/mm/yyyy), HomeTeam, AwayTeam, FTHG, FTAG

    df <- df %>%
      mutate(
        # Parse date (football-data uses d/m/Y)
        Date = suppressWarnings(lubridate::dmy(Date)),

        # Standardize columns your app expects
        Home = if ("Home" %in% names(df)) Home else HomeTeam,
        Away = if ("Away" %in% names(df)) Away else AwayTeam,

        HomeGoals = if ("HomeGoals" %in% names(df)) HomeGoals else as.numeric(FTHG),
        AwayGoals = if ("AwayGoals" %in% names(df)) AwayGoals else as.numeric(FTAG),

        # League name your UI uses
        Competition_Name = case_when(
          Div == "E0"  ~ "Premier League",
          Div == "D1"  ~ "Fußball-Bundesliga",
          Div == "SP1" ~ "La Liga",
          Div == "F1"  ~ "Ligue 1",
          Div == "I1"  ~ "Serie A",
          TRUE ~ as.character(Div)
        )
      )
  } else {
    # Old schema: just make sure Date is Date
    df <- df %>% mutate(Date = as.Date(Date))
  }

  # Return a consistent minimum set + keep extras if you want them later
  df %>%
    select(Competition_Name, Date, Home, Away, HomeGoals, AwayGoals, everything())
}

```


``` {r}

## Load data
team_mapping <- read_csv("https://raw.githubusercontent.com/steodose/Club-Soccer-Forecasts/main/team_mapping.csv")

# fix Nottingham Forest in 538 team mapping data
team_mapping2 <- team_mapping %>%
    mutate(team_538 = case_when(
        team == "Nott'm Forest" ~ "Nottingham Forest",
        TRUE ~ team_538
    ))


# updated team mappings for 2024-25 fixtures
team_mapping_2024 <- read_csv('https://raw.githubusercontent.com/steodose/club-elo/main/all_leagues_team_mapping.csv')

# rename
team_mapping_2024 <- team_mapping_2024 %>%
  mutate(team = case_when(
        team == "Tottenham Hotspur" ~ "Tottenham",
        team == "Newcastle" ~ "Newcastle Utd",
        team == "Manchester United" ~ "Manchester Utd",
        team == "West Ham United" ~ "West Ham",
        team == "Brighton and Hove Albion" ~ "Brighton",
        team == "AFC Bournemouth" ~ "Bournemouth",
        team == "Wolverhampton" ~ "Wolves",
        team == "Nottingham Forest" ~ "Nott'ham Forest",
        team == "Girona FC" ~ "Girona",
        team == "Atletico Madrid" ~ "Atlético Madrid",
        team == "Athletic Bilbao" ~ "Athletic Club",
        team == "Deportivo Alaves" ~ "Alavés",
        team == "Sevilla FC" ~ "Sevilla",
        team == "Real Betis" ~ "Betis",
        team == "Internazionale" ~ "Inter",
        team == "AC Milan" ~ "Milan",
        team == "Bayer Leverkusen" ~ "Leverkusen",
        team == "VfB Stuttgart" ~ "Stuttgart",
        team == "Borussia Dortmund" ~ "Dortmund",
        team == "Eintracht Frankfurt" ~ "Eint Frankfurt",
        team == "TSG Hoffenheim" ~ "Hoffenheim",
        team == "SC Freiburg" ~ "Freiburg",
        team == "FC Augsburg" ~ "Augsburg",
        team == "VfL Wolfsburg" ~ "Wolfsburg",
        team == "Mainz 05" ~ "Mainz",
        team == "Borussia Monchengladbach" ~ "Gladbach",
        team == "Bochum" ~ "VfLBochum",
        team == "Paris Saint-Germain" ~ "Paris S-G",
        team == "AS Momaco" ~ "Monaco",
        team == "AS Roma" ~ "Roma",
        TRUE ~ team
    ))



# --- PROD (GitHub) ---
matches_raw <- read_csv(
  "https://raw.githubusercontent.com/steodose/club-elo/main/matches_all_leagues.csv",
  show_col_types = FALSE
)
matches <- clean_matches(matches_raw)

# --- LOCAL TEST (uncomment when needed) ---
# matches_raw <- read_csv("matches_all_leagues.csv", show_col_types = FALSE)
# matches <- clean_matches(matches_raw)

# read in latest Elo ratings
latest_ratings <- read_csv('https://raw.githubusercontent.com/steodose/club-elo/main/latest_ratings.csv')

# read in latest simulations
# sim_rank_dist <- read_csv('https://raw.githubusercontent.com/steodose/club-elo/main/simulations_rank_dist.csv', show_col_types = FALSE)
# sim_summary   <- read_csv('https://raw.githubusercontent.com/steodose/club-elo/main/simulations_summary.csv', show_col_types = FALSE)

sim_rank_dist <- read_csv("simulations_rank_dist.csv")
sim_summary   <- read_csv("simulations_summary.csv")

```

Inputs {.sidebar}
-----------------------------------------------------------------------

``` {r}

selectInput(
  "leagueInput",
  label = h3("Select League"),
  choices = sort(unique(matches$Competition_Name)),
  selected = "Premier League",
  multiple = FALSE
)

```


Row
-----------------------------------------------------------------------

### Country {.value-box data-width=50}

```{r}

# make dataframe with country names
country_leagues <- data.frame(country = c("England", "Germany", "Spain", "France", "Italy"),
                        league = c("Premier League", "Fußball-Bundesliga", "La Liga", "Ligue 1", "Serie A")
)

# add flag mappings
country_leagues <- data.frame(
  country = c("England", "Germany", "Spain", "France", "Italy"),
  league  = c("Premier League", "Fußball-Bundesliga", "La Liga", "Ligue 1", "Serie A"),
  flag_url = c(
    "https://flagcdn.com/w40/gb-eng.png",  # England
    "https://flagcdn.com/w40/de.png",
    "https://flagcdn.com/w40/es.png",
    "https://flagcdn.com/w40/fr.png",
    "https://flagcdn.com/w40/it.png"
  )
)


country <- reactive({
  country_leagues %>% 
  filter(league == input$leagueInput) %>% 
  pull(country)
})


# Put valueBox() inside renderValueBox({})
renderValueBox({

  df <- country_leagues %>%
    filter(league == input$leagueInput)

  valueBox(
    value = HTML(
      paste0(
        '<img src="', df$flag_url, '" height="24" style="margin-right:8px;">',
        df$country
      )
    ),
    caption = "Country",
    color = "#013369",
    icon = NULL
  )
})

```


### Data Updated {.value-box data-width=50}

```{r}

update_date <- reactive({
  matches %>%
    filter(Competition_Name == input$leagueInput) %>%
    filter(!is.na(HomeGoals), !is.na(AwayGoals)) %>%   # only completed matches
    summarise(latest_match = max(Date, na.rm = TRUE)) %>%
    pull(latest_match)
})


# Put valueBox() inside renderValueBox({})
renderValueBox({
  valueBox(
    value = format(update_date(), "%B %d, %Y"),
    caption = "Matches Last Updated",
    color = "#013369",
    icon = "fa-calendar"
  )
})

```

Row {.tabset data-height=650}
-----------------------------------------------------------------------

### League Tables
    
```{r}

league_matches <- reactive({
  matches %>%
  filter(Competition_Name == input$leagueInput) %>%
  mutate(result = HomeGoals - AwayGoals) %>%
    select(Home, Away, HomeGoals, AwayGoals, result) %>%
    pivot_longer(Home:Away, names_to = "home_away", values_to = "team") %>%
    mutate(
        result = ifelse(home_away == "Home", result, -result),
        win = ifelse(result == 0, 0.5, ifelse(result > 0, 1, 0))
    ) %>%
    select(team, HomeGoals, AwayGoals, win, result) %>%
    drop_na() %>%
    group_by(team) %>%
    summarise(
        Wins = length(win[win == 1]),
        Losses = length(win[win == 0]),
        Draws = length(win[win == 0.5]),
        MP = sum(Wins, Losses, Draws),
        Points = (Wins * 3) + (Draws * 1),
        win_perc = (100 * Points / (MP * 3)),
        GD = sum(result),
        form = list(win), .groups = "drop"
    ) %>%
    left_join(team_mapping_2024, by = c("team" = "team")) %>%
    select(url_logo_espn, team, Points, MP, Wins, Draws, Losses, GD, win_perc, form) %>%
    arrange(desc(Points), desc(GD)) %>%
    ungroup() %>%
    mutate(Rank = row_number()) %>%
    relocate(Rank) %>%
    rename(Team = team) %>%
    mutate(list_data = list(c(Wins, Draws, Losses))) %>%
    gather(attr_num, list_data, c(Wins, Draws, Losses)) %>%
    group_by_at(vars(-attr_num, -list_data)) %>%
    summarise(list_data = list(list_data)) %>%
    ungroup()
})



# Make current standings table using gt package
render_gt({
league_matches() %>%
    gt() %>%
    # Relabel columns
    cols_label(
        url_logo_espn = "",
        win_perc = "Win %",
        form = "Form"
    ) %>%
    text_transform(
        locations = cells_body(vars(url_logo_espn)),
        fn = function(x) {
            web_image(url = x,
                      height = px(30))
        }
    ) %>%
    data_color(columns = 4,
               colors = scales::col_numeric(
                   palette = c("white", "#3fc1c9"),
                   domain = NULL)
    ) %>%
    #gt_theme_538() %>%
    tab_style(
        style = cell_borders(sides = "bottom", color = "black", weight = px(1)),
        locations = cells_body(rows = 4)
    ) %>%
    tab_style(
        style = cell_borders(sides = "bottom", color = "black", weight = px(1)),
        locations = cells_body(rows = 17)
    ) %>%
    tab_style(
        style = list(
            cell_text(color = "red")
        ),
        locations = cells_body(
            columns = vars(GD),
            rows = GD <= 0
        )
    ) %>%
    tab_style(
        style = list(
            cell_text(color = "blue")
        ),
        locations = cells_body(
            columns = vars(GD),
            rows = GD > 0
        )
    ) %>%
    # tab_style(
    #     style = list(
    #         cell_text(color = "red")
    #     ),
    #     locations = cells_body(
    #         columns = vars(xGD),
    #         rows = xGD <= 0
    #     )
    # ) %>%
    # tab_style(
    #     style = list(
    #         cell_text(color = "blue")
    #     ),
    #     locations = cells_body(
    #         columns = vars(xGD),
    #         rows = xGD > 0
    #     )
    # ) %>%
    # fmt_number(
    #     columns = vars(xGD),
    #     decimals = 1
    # ) %>%
    cols_align(align = "left",
               columns = 1) %>%
    tab_options(column_labels.font.weight = "bold") %>%
    tab_header(
  title = md(glue("**2025-26 {input$leagueInput} Table**")),
  subtitle = glue("Thru matches played as of date mentioned above.")) %>%
    tab_source_note(
        source_note = md("DATA: FBREF.com")) %>%
    gt_plt_bar_pct(column = win_perc, scaled = TRUE, fill = "navy", background = "gray") %>%
    gt_plt_winloss(form, max_wins = 30) %>%
    gt_plt_bar_stack(list_data, width = 60,
                     labels = c("  WINS  ", "  DRAWS  ", "  LOSSES  "),
                     palette= c("#ff4343", "#bfbfbf", "#0a1c2b")) %>%
    tab_footnote(
        footnote = "Share of total available points captured. Win = 3 points, Draw = 1 point, Loss = 0 points.",
        locations = cells_column_labels(vars(win_perc))
    ) 
  # %>%
    # tab_footnote(
    #     footnote = "Expected Goals (xG) is the probability a shot will result in a goal based on the characteristics of that shot. Measures  quality of chances created.",
    #     locations = cells_column_labels(vars(xGD))
    # )

})

```



```{css}
    .chart-shim {
      overflow: auto;
    }
```

<!-- ### All Team Ratings -->

<!-- **Big 5 European Team Ratings** -->

<!-- 2025-26 ratings coming soon... -->

<!-- ``` {r} -->

<!-- # Reactable table setup -->
<!-- latest_ratings_table <- reactive({ -->
<!--   latest_ratings %>% -->
<!--   left_join(team_mapping_2024) %>% -->
<!--   select(url_logo_espn, team, league, country, rating) %>% -->
<!--   arrange(desc(rating)) %>% -->
<!--   mutate(rank = row_number() -->
<!--          ) %>% -->
<!--   select(rank, url_logo_espn, team,league, country, rating) -->
<!--     }) -->


<!-- # render reactable table -->
<!-- renderReactable({ -->
<!-- reactable(latest_ratings_table(), -->
<!--           #theme = theme_538, -->
<!--                   showSortIcon = TRUE, -->
<!--                   searchable = TRUE, -->
<!--                   language = reactableLang( -->
<!--                       searchPlaceholder = "SEARCH FOR A TEAM..."), -->
<!--                   defaultPageSize = 100, -->
<!--                   columns = list( -->
<!--                     rank = colDef(name = "Rank", -->
<!--                                        align = "center"), -->
<!--                       team = colDef(name = "Team", -->
<!--                                     minWidth = 140, -->
<!--                                     align = "left"), -->
<!--                     league = colDef(name = "League", -->
<!--                                     minWidth = 140, -->
<!--                                     align = "left"), -->
<!--                     country = colDef(name = "Country", -->
<!--                                     minWidth = 140, -->
<!--                                     align = "left"), -->
<!--                     rating = colDef(name = "Rating", -->
<!--                                       align = "right", -->
<!--                                       style = color_scales(latest_ratings, colors = paletteer::paletteer_d( -->
<!--                                          palette = "ggsci::amber_material" -->
<!--                                      ) -->
<!--                                      ), -->
<!--                                       format =  colFormat(digits = 1)), -->

<!--                       ### add logos using embed_img() -->
<!--                       url_logo_espn = colDef( -->
<!--                           name = "", -->
<!--                           maxWidth = 70, -->
<!--                           align = "center", -->
<!--                           cell = embed_img(height = "25", width = "28") -->
<!--                       ) -->
<!--                       ), -->
<!--                   pagination = FALSE, -->
<!--                   compact = TRUE,  -->
<!--                   borderless = FALSE,  -->
<!--                   striped = FALSE, -->
<!--                   fullWidth = FALSE,  -->
<!--                   defaultColDef = colDef(align = "center", minWidth = 95) -->
<!--           ) -->
<!-- }) -->

<!-- ``` -->

### Forecasts

2025-26 forecasts for all leagues coming soon...

```{r}

# 1) Build the league-specific sims table reactively
forecast_table <- reactive({

  sim_league <- sim_rank_dist %>%
    filter(Competition_Name == input$leagueInput) %>%
    mutate(Placement = as.character(Rank)) %>%
    select(Team, Placement, Freq = prob)

  validate(
    need(nrow(sim_league) > 0, "No simulations found for this league yet.")
  )

  # Wide table: Team + rank probability columns ("1","2",...)
  league_table3 <- sim_league %>%
      pivot_wider(names_from  = Placement,
                  values_from = Freq,
                  values_fill = 0
                  )

rank_cols <- setdiff(names(league_table3), "Team")
rank_cols_num <- suppressWarnings(as.integer(rank_cols))
rank_cols <- rank_cols[order(rank_cols_num)]

league_table3 <- league_table3 %>%
  mutate(across(all_of(rank_cols), ~ as.numeric(.x)))

  # Identify rank columns from the wide table (everything except Team)
  rank_cols <- setdiff(names(league_table3), "Team")
  rank_cols_num <- suppressWarnings(as.integer(rank_cols))
  rank_cols <- rank_cols[order(rank_cols_num)]

  validate(
    need(length(rank_cols) > 0, "No rank probability columns found (expected ranks like '1','2',...).")
  )

  # FORCE rank columns numeric (fixes rescale() on character)
  league_table3 <- league_table3 %>%
    mutate(across(all_of(rank_cols), ~ as.numeric(.x)))

  # Add logos AFTER numeric coercion
  league_table_gt <- league_table3 %>%
    left_join(team_mapping_2024, by = c("Team" = "team")) %>%
    select(url_logo_espn, Team, all_of(rank_cols)) %>%
    rename(logo = url_logo_espn)

  # Order by title odds, then runner-up odds (or next best)
  league_table_gt3 <- league_table_gt %>%
    arrange(
      desc(.data[[rank_cols[1]]]),
      desc(.data[[rank_cols[min(2, length(rank_cols))]]])
    ) %>%
    mutate(Rank = row_number()) %>%
    relocate(Rank)

  list(
    data = league_table_gt3,
    rank_cols = rank_cols
  )
})

# 2) Render GT table (reactive consumer)
render_gt({
  obj <- forecast_table()
  league_table_gt3 <- obj$data
  rank_cols <- obj$rank_cols

  league_table_gt3 %>%
    gt() %>%
    cols_label(Rank = "", logo = "", Team = "") %>%
    tab_header(
      title = md(glue("**{input$leagueInput} Simulations**")),
      subtitle = paste0("2025-26 Season | Updated ", format(Sys.Date(), format = "%B %d, %Y"))
    ) %>%
    text_transform(
      locations = cells_body(vars(logo)),
      fn = function(x) web_image(url = x, height = px(30))
    ) %>%
    fmt_percent(columns = all_of(rank_cols), decimals = 1) %>%
    data_color(
      columns = all_of(rank_cols),
      colors = scales::col_numeric(
        palette = c("white", "#ffbf00", "#ff7a00", "#b30000"),
        domain = c(0, 1),
        na.color = "white"
      )
    ) %>%
    tab_options(
      column_labels.background.color = "white",
      table.border.top.width = px(3),
      table.border.top.color = "white",
      table.border.bottom.color = "white",
      table.border.bottom.width = px(3),
      column_labels.border.top.width = px(3),
      column_labels.border.top.color = "white",
      column_labels.border.bottom.width = px(3),
      column_labels.border.bottom.color = "black",
      data_row.padding = px(3),
      source_notes.font.size = 12,
      table.font.size = 16,
      heading.align = "left"
    ) %>%
    tab_source_note(source_note = md("Table: @steodosescu | Data: football-data.co.uk"))
})


```


### About

**Big 5 European Leagues App**

This dashboard looks at the current tables and statistics for the Big 5 European leagues. All data is sourced from Football Reference (FBREF.com) unless noted otherwise. App created using R Shiny by [Stephan Teodosescu](https://stephanteodosescu.com).