---
title: "Big 5 Leagues"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: cosmo
    social: [ "twitter", "linkedin", "menu" ]
---

<style>                     
.navbar {
  background-color:black;
  border-color:white;
}
.navbar-brand {
color:white!important;
}
</style>   

<style type="text/css">

.chart-title {  /* chart_title  */
   font-size: 15px

</style>


```{r global, include = FALSE}
library(flexdashboard)
library(shiny)
library(reactable)
library(reactablefmtr)
library(htmltools)
library(worldfootballR)
library(tidyverse)
library(rvest)
library(extrafont)
library(gt)
library(gtExtras)
library(ggimage) 
library(glue)
library(ggtext)
library(scales)
library(svglite)


## Setup steps

# Optional but makes R prefer not to display numbers in scientific notation
options(scipen = 9999)

# Custom ggplot theme (inspired by Owen Phillips at the F5 substack blog)
theme_custom <- function () { 
  theme_minimal(base_size=11, base_family="Chivo") %+rbundace% 
    theme(
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = 'transparent', color = "transparent")
    )
}

# Define 538 table theme for Reactable table(s) below
theme_538 <- function() {
    reactable::reactableTheme(
        searchInputStyle = list(width = "31%", backgroundColor = "#F9F9F9"),
        style = list(
            fontFamily = "Chivo"
        ),
        headerStyle = list(
            "&:hover[aria-sort]" = list(
                background = "hsl(0, 0%, 80%)"),
            "&[aria-sort='ascending'], &[aria-sort='descending']" = list(
                background = "#555",
                color = "#FFF"
            ),
            borderColor = "#333"
        ),
        borderColor = "#CDCDCD"
    )
}


##### Custom gt table themes for graphics. Inspired by Tom Mock's excellent blog posts #####
gt_theme_538 <- function(data,...) {
    data %>%
        # Add team logos w/ web_image
        text_transform(
            locations = cells_body(
                columns = flag_logo
            ),
            fn = function(x) {
                web_image(
                    url = x,
                    height = 25
                )
            }
        ) %>%
        # Relabel columns
        cols_label(
            flag_logo = ""
        ) %>%
        opt_all_caps()  %>%
        opt_table_font(
            font = list(
                google_font("Chivo"),
                default_fonts()
            )
        ) %>%
        tab_style(
            style = cell_borders(
                sides = "bottom", color = "transparent", weight = px(2)
            ),
            locations = cells_body(
                columns = TRUE,
                # This is a relatively sneaky way of changing the bottom border
                # Regardless of data size
                rows = nrow(data$`_data`)
            )
        )  %>% 
        tab_options(
            column_labels.background.color = "white",
            table.border.top.width = px(3),
            table.border.top.color = "transparent",
            table.border.bottom.color = "transparent",
            table.border.bottom.width = px(3),
            column_labels.border.top.width = px(3),
            column_labels.border.top.color = "transparent",
            column_labels.border.bottom.width = px(3),
            column_labels.border.bottom.color = "black",
            data_row.padding = px(3),
            source_notes.font.size = 12,
            table.font.size = 16,
            heading.align = "left",
            ...
        ) 
}

```


``` {r}

## Load data
team_mapping <- read_csv("https://raw.githubusercontent.com/steodose/Club-Soccer-Forecasts/main/team_mapping.csv")

# fix Nottingham Forest in 538 team mapping data
team_mapping2 <- team_mapping %>%
    mutate(team_538 = case_when(
        team == "Nott'm Forest" ~ "Nottingham Forest",
        TRUE ~ team_538
    ))


# updated team mappings for 2024-25 fixtures
team_mapping_2024 <- read_csv('https://raw.githubusercontent.com/steodose/club-elo/main/all_leagues_team_mapping.csv')

# rename
team_mapping_2024 <- team_mapping_2024 %>%
  mutate(team = case_when(
        team == "Tottenham Hotspur" ~ "Tottenham",
        team == "Newcastle" ~ "Newcastle Utd",
        team == "Manchester United" ~ "Manchester Utd",
        team == "West Ham United" ~ "West Ham",
        team == "Brighton and Hove Albion" ~ "Brighton",
        team == "AFC Bournemouth" ~ "Bournemouth",
        team == "Wolverhampton" ~ "Wolves",
        team == "Nottingham Forest" ~ "Nott'ham Forest",
        team == "Girona FC" ~ "Girona",
        team == "Atletico Madrid" ~ "Atlético Madrid",
        team == "Athletic Bilbao" ~ "Athletic Club",
        team == "Deportivo Alaves" ~ "Alavés",
        team == "Sevilla FC" ~ "Sevilla",
        team == "Real Betis" ~ "Betis",
        team == "Internazionale" ~ "Inter",
        team == "AC Milan" ~ "Milan",
        team == "Bayer Leverkusen" ~ "Leverkusen",
        team == "VfB Stuttgart" ~ "Stuttgart",
        team == "Borussia Dortmund" ~ "Dortmund",
        team == "Eintracht Frankfurt" ~ "Eint Frankfurt",
        team == "TSG Hoffenheim" ~ "Hoffenheim",
        team == "SC Freiburg" ~ "Freiburg",
        team == "FC Augsburg" ~ "Augsburg",
        team == "VfL Wolfsburg" ~ "Wolfsburg",
        team == "Mainz 05" ~ "Mainz",
        team == "Borussia Monchengladbach" ~ "Gladbach",
        team == "Bochum" ~ "VfLBochum",
        TRUE ~ team
    ))


# matches <- read_csv("https://projects.fivethirtyeight.com/soccer-api/club/spi_matches_latest.csv") %>%
#   filter(league == 'Barclays Premier League' | league == 'German Bundesliga' | league == 'Spanish Primera Division' |  league == 'French Ligue 1' | league == 'Italy Serie A',
#            season == 2022)

# Extra code to get unique teams
# teams1 <- matches %>%
#   distinct(team1)
# 
# teams2 <- matches %>%
#   distinct(team2)
# 
# teams <- bind_rows(teams1, teams2) %>%
#   select(team1) %>%
#   rename(team = team1) %>%
#   distinct(team) %>%
#   drop_na
# 
# write_csv(teams, "big_5_team_mapping.csv")


## Load data V2 from football data

# matches_eng <- read_csv("https://www.football-data.co.uk/mmz4281/2324/E0.csv")
# matches_esp <- read_csv("https://www.football-data.co.uk/mmz4281/2324/SP1.csv")
# matches_ger <- read_csv("https://www.football-data.co.uk/mmz4281/2324/D1.csv")
# matches_fra <- read_csv("https://www.football-data.co.uk/mmz4281/2324/F1.csv")
# matches_ita <- read_csv("https://www.football-data.co.uk/mmz4281/2324/I1.csv")

## Load data V3 from worldfootballr
# matches_eng <- worldfootballR::fb_match_results("ENG", "M", season_end_year = 2024)
# matches_esp <- worldfootballR::fb_match_results("ESP", "M", season_end_year = 2024)
# matches_ger <- worldfootballR::fb_match_results("GER", "M", season_end_year = 2024)
# matches_fra <- worldfootballR::fb_match_results("FRA", "M", season_end_year = 2024)
# matches_ita <- worldfootballR::fb_match_results("ITA", "M", season_end_year = 2024)
# 
# 
# # join all leagues
# matches <- bind_rows(matches_eng, matches_esp, matches_ger, matches_fra, matches_ita)
# 
# matches <- matches %>%
#    mutate(Div = case_when(
#         Div == "E0" ~ "EPL",
#         Div == "SP1" ~ "La Liga",
#         Div == "D1" ~ "Bundesliga",
#         Div == "F1" ~ "Ligue 1",
#         Div == "I1" ~ "Serie A",
#         TRUE ~ Div
#     ))


# updated matches for 2024-25
matches <- read_csv('https://raw.githubusercontent.com/steodose/club-elo/main/matches_all_leagues.csv')

# read in latest Elo ratings
latest_ratings <- read_csv('https://raw.githubusercontent.com/steodose/club-elo/main/latest_ratings.csv')

```

Inputs {.sidebar}
-----------------------------------------------------------------------

``` {r}

 selectInput("leagueInput", label = h3("Select League"), 
    choices = unique(matches$Competition_Name),
    selected="EPL", multiple = FALSE)

```


Row
-----------------------------------------------------------------------

### Country {.value-box data-width=50}

```{r}

# make dataframe with country names
country_leagues <- data.frame(country = c("England", "Germany", "Spain", "France", "Italy"),
                        league = c("Premier League", "Fußball-Bundesliga", "La Liga", "Ligue 1", "Serie A")
)

country <- reactive({
  country_leagues %>% 
  filter(league == input$leagueInput) %>% 
  pull(country)
})


# Put valueBox() inside renderValueBox({})
renderValueBox({
  valueBox(value = country(), 
           caption = "Country",
           color = "#013369",
           icon = "fa-globe")
})

```


### Data Updated {.value-box data-width=50}

```{r}

update_date <- reactive({
  matches %>% 
  filter(Competition_Name == input$leagueInput) %>% 
    select(-Notes, -Round) %>% #remove this field so the next line works properly
    na.omit() %>%
  summarise(latest_match = max(Date, na.rm = TRUE))
})

# Put valueBox() inside renderValueBox({})
renderValueBox({
  valueBox(value = update_date(), 
           caption = "Matches Last Updated",
           color = "#013369",
           icon = "fa-calendar")
})

```

Row {.tabset data-height=650}
-----------------------------------------------------------------------

### League Tables
    
```{r}

league_matches <- reactive({
  matches %>%
  filter(Competition_Name == input$leagueInput) %>%
  mutate(result = HomeGoals - AwayGoals) %>%
    select(Home, Away, HomeGoals, AwayGoals, result) %>%
    pivot_longer(Home:Away, names_to = "home_away", values_to = "team") %>%
    mutate(
        result = ifelse(home_away == "Home", result, -result),
        win = ifelse(result == 0, 0.5, ifelse(result > 0, 1, 0))
    ) %>%
    select(team, HomeGoals, AwayGoals, win, result) %>%
    drop_na() %>%
    group_by(team) %>%
    summarise(
        Wins = length(win[win == 1]),
        Losses = length(win[win == 0]),
        Draws = length(win[win == 0.5]),
        MP = sum(Wins, Losses, Draws),
        Points = (Wins * 3) + (Draws * 1),
        win_perc = (100 * Points / (MP * 3)),
        GD = sum(result),
        form = list(win), .groups = "drop"
    ) %>%
    left_join(team_mapping_2024, by = c("team" = "team")) %>%
    select(url_logo_espn, team, Points, MP, Wins, Draws, Losses, GD, win_perc, form) %>%
    arrange(desc(Points), desc(GD)) %>%
    ungroup() %>%
    mutate(Rank = row_number()) %>%
    relocate(Rank) %>%
    rename(Team = team) %>%
    mutate(list_data = list(c(Wins, Draws, Losses))) %>%
    gather(attr_num, list_data, c(Wins, Draws, Losses)) %>%
    group_by_at(vars(-attr_num, -list_data)) %>%
    summarise(list_data = list(list_data)) %>%
    ungroup()
})



# Make current standings table using gt package
render_gt({
league_matches() %>%
    gt() %>%
    # Relabel columns
    cols_label(
        url_logo_espn = "",
        win_perc = "Win %",
        form = "Form"
    ) %>%
    text_transform(
        locations = cells_body(vars(url_logo_espn)),
        fn = function(x) {
            web_image(url = x,
                      height = px(30))
        }
    ) %>%
    data_color(columns = 4,
               colors = scales::col_numeric(
                   palette = c("white", "#3fc1c9"),
                   domain = NULL)
    ) %>%
    #gt_theme_538() %>%
    tab_style(
        style = cell_borders(sides = "bottom", color = "black", weight = px(1)),
        locations = cells_body(rows = 4)
    ) %>%
    tab_style(
        style = cell_borders(sides = "bottom", color = "black", weight = px(1)),
        locations = cells_body(rows = 17)
    ) %>%
    tab_style(
        style = list(
            cell_text(color = "red")
        ),
        locations = cells_body(
            columns = vars(GD),
            rows = GD <= 0
        )
    ) %>%
    tab_style(
        style = list(
            cell_text(color = "blue")
        ),
        locations = cells_body(
            columns = vars(GD),
            rows = GD > 0
        )
    ) %>%
    # tab_style(
    #     style = list(
    #         cell_text(color = "red")
    #     ),
    #     locations = cells_body(
    #         columns = vars(xGD),
    #         rows = xGD <= 0
    #     )
    # ) %>%
    # tab_style(
    #     style = list(
    #         cell_text(color = "blue")
    #     ),
    #     locations = cells_body(
    #         columns = vars(xGD),
    #         rows = xGD > 0
    #     )
    # ) %>%
    # fmt_number(
    #     columns = vars(xGD),
    #     decimals = 1
    # ) %>%
    cols_align(align = "left",
               columns = 1) %>%
    tab_options(column_labels.font.weight = "bold") %>%
    tab_header(title = md("**2024-25 League Table**"),
               subtitle = glue("Thru matches played as of date mentioned above.")) %>%
    tab_source_note(
        source_note = md("DATA: FBREF.com")) %>%
    gt_plt_bar_pct(column = win_perc, scaled = TRUE, fill = "navy", background = "gray") %>%
    gt_plt_winloss(form, max_wins = 30) %>%
    gt_plt_bar_stack(list_data, width = 60,
                     labels = c("  WINS  ", "  DRAWS  ", "  LOSSES  "),
                     palette= c("#ff4343", "#bfbfbf", "#0a1c2b")) %>%
    tab_footnote(
        footnote = "Share of total available points captured. Win = 3 points, Draw = 1 point, Loss = 0 points.",
        locations = cells_column_labels(vars(win_perc))
    ) 
  # %>%
    # tab_footnote(
    #     footnote = "Expected Goals (xG) is the probability a shot will result in a goal based on the characteristics of that shot. Measures  quality of chances created.",
    #     locations = cells_column_labels(vars(xGD))
    # )

})

```



```{css}
    .chart-shim {
      overflow: auto;
    }
```

### All Team Ratings

**Big 5 European Team Ratings**

2024-25 ratings coming soon...

``` {r}

# Reactable table setup
latest_ratings_table <- reactive({
  latest_ratings %>%
  left_join(team_mapping_2024) %>%
  select(url_logo_espn, team, league, country, rating) %>%
  arrange(desc(rating)) %>%
  mutate(rank = row_number()
         ) %>%
  select(rank, url_logo_espn, team,league, country, rating)
    })


# render reactable table
renderReactable({
reactable(latest_ratings_table(),
          #theme = theme_538,
                  showSortIcon = TRUE,
                  searchable = TRUE,
                  language = reactableLang(
                      searchPlaceholder = "SEARCH FOR A TEAM..."),
                  defaultPageSize = 100,
                  columns = list(
                    rank = colDef(name = "Rank",
                                       align = "center"),
                      team = colDef(name = "Team",
                                    minWidth = 140,
                                    align = "left"),
                    league = colDef(name = "League",
                                    minWidth = 140,
                                    align = "left"),
                    country = colDef(name = "Country",
                                    minWidth = 140,
                                    align = "left"),
                    rating = colDef(name = "Rating",
                                      align = "right",
                                      style = color_scales(latest_ratings, colors = paletteer::paletteer_d(
                                         palette = "ggsci::amber_material"
                                     )
                                     ),
                                      format =  colFormat(digits = 1)),
                      
                      ### add logos using embed_img()
                      url_logo_espn = colDef(
                          name = "",
                          maxWidth = 70,
                          align = "center",
                          cell = embed_img(height = "25", width = "28")
                      )
                      ),
                  pagination = FALSE,
                  compact = TRUE, 
                  borderless = FALSE, 
                  striped = FALSE,
                  fullWidth = FALSE, 
                  defaultColDef = colDef(align = "center", minWidth = 95)
          )
})

```



### About

**Big 5 European Leagues App**

This dashboard looks at the current tables and statistics for the Big 5 European leagues. All data is sourced from Football Reference (FBREF.com) unless noted otherwise. App created using R Shiny by [Stephan Teodosescu](https://stephanteodosescu.com).